\section{Introduction}

%Speed Up necessity

In a world reliant on computer systems, the correctness of those systems is vital. Indeed, simple programming errors can lead to major incidents; examples of these include an automated trader losing \$460 million \cite{KnightCapital} and the inaugural Ariane 5 flight breaking up after launch due to an overflow error \cite{Flight501Failure}. 
%There are also various safety concerns, including data loss and critical system outages; these need to be avoided

In order to achieve better performance, concurrency can often be introduced to improve the performance of programs or systems, especially those with semi-independent tasks or components.
Concurrent systems, be this on a single computer or distributed across a network, achieve these performance improvements at the cost of additional complexity; the design now needs to consider the interactions between threads. 
Each of possible interactions between threads could potentially lead to a \emph{race condition} in a poorly designed systems. A race condition is where two non-independent actions can occur in an order which produces an incorrect or unwanted outcome. 
%The most famous example of this was in the Therac-25 radiation therapy machine which lead to the deaths of three patients due to overexposure as a result of a race condition in the machine's software where the high radiation beam was selected and quickly changed to the lower radiation beam\cite{BugSnag}. 
Race conditions can be very damaging in practice: the Therac-25 radiation therapy machine killed three of its patients by radiation overexposure as a result of a race condition \cite{BugSnag}. This was caused by the operator quickly changing from the high radiation beam mode to select the lower radiation beam instead; the race condition resulted in the machine erroneously still using the high radiation beam. This highlights the importance of thorough system validation; had more through system verification been completed, these deaths would have likely been avoided \cite{AGift}.


%The Therac-25 radiation therapy machine had an example of this, where an operator quickly changing from the high radiation beam to select the lower radiation beam would still result in the machine still using the high radiation beam because of a race condition. This error resulted in the deaths of three patients\cite{BugSnag} and, had more through system validation been completed, these deaths would have likely been avoided\cite{AGift}.

There are two main approaches to developing correct software: testing and verification. 
Testing can only establish the presence of a bug, not the absence of any; this can be somewhat addressed by writing exhaustive tests to cover every possible edge case, however this imposes severe restrictions on the complexity of such systems. Though thorough unit testing can be effective in reducing software bugs \cite{MicroTest}, this form of testing is significantly less effective in concurrent contexts. This is due to the inherent nondeterminism of event scheduling; this make accuractely testing all possible interactions between threads a significant challenge.

%Writing exhaustive tests is near impossible for sufficiently complex concurrent systems. 
%This is predominantly due to the sheer number of different interactions between independent threads: considering all possible interactions and testing them effectively are both challenging tasks.
\emph{Linearization} testing is an effective alternative approach.
This treats each operation as having occured (been \emph{linearized}) at some point in time between the start and finish of the operation. It then checks that performing the operations sequentially in the same order returns the same results.
This also relies on the nondeterministic testing of edge cases so is also not exhaustive either \cite{LoweLin}. %revisit

By contrast, formal verification can be used to show that systems satisfy some desired properties \cite{PrinciplesOfModelChecking}. This, however, is a complex process and it is often impossible to model check the complete behaviour of a large system simply due to the size of the resulting state space; we instead focus on modelling the concurrent interactions between threads via a variety of synchronisation objects. Synchronisation objects are used to control the potential interaction between threads, simplifying the design of concurrent systems and helping to achieve race conditions.

There are two main benefits of this. Firstly, we prove that the synchronisation objects function correctly; debugging concurrent programs is a complex task and bugs due to issues with synchronisation objects can often be particularly challenging to resolve. The second benefit is that we can significantly optimise the model checking of systems that use these primitives. 
If we can prove that some synchronisation object fulfils a simpler specification which has the same properties, we can then use the specification to simplify future model checks but wihtout changing the system behaviour.%revisit
%We prove that the synchronisation objects fulfil a specification with the same properties but without all of the internal workings. We can therefore optimise model checking of whole systems by replacing the original synchronisation objects with their respective specification process without changing system behaviour. 
This allows us to significantly reduce the possible state space of the system, making formal verification of complex systems significantly more feasible.
%; if the interactions between threads behave as expected then system validation can be reduced to checking the behaviour of the individual, sequential, threads.

%Focusing on these interactions allows us to decrease the size of a system to the extent where it is now feasible to model check these behaviours. 

We choose to use the process algebra Communicating Sequential Processes (CSP) \cite{RoscoeUCS} as our tool for modelling these interactions and the accompanying FDR model checker \cite{GibsonFDR}. CSP is a language for describing processes that can interact both with their environment and other parallel processes. Since CSP is covered in the Part B Concurrency course \cite{Concurrency}, we assume familiarity with it (as described in \cite{RoscoeUCS}) throughout.


%As a result, we can focus our modelling on the concurrent interactions between threads, be this directly via message passing or through concurrent datatypes or primitives. 
%If the interactions between threads behave as expected then system validation can instead be reduced to checking the behaviour of the individual sequential threads. Focussing on these interactions decreases the size of the system to the extent where it is now feasible to model check the concurrent behaviours. We choose to use the process algebra Communicating Sequential Processes (CSP)\cite{Roscoe} as our tool for modelling these interactions

\subsection{Contributions}

The contributions of this project are as follows:

\begin{itemize}
  \item We examine a variety of locks from \cite{CADS} and a number of desirable properties of locks, using CSP refinement checks to determine which properties each lock satisfes. We also discuss the feasibility of modelling infinite liveness properties using CSP, presenting a novel proof that starvation-freedom cannot be captured in CSP before considering alternative approaches.%revisit - last bit
  \item We model both the standard JVM monitor and the Scala Concurrency Library (SCL) \cite{CP} implementation of a monitor, proving that the SCL monitor provides the same mutual exclusion and correctness properties as the JVM monitor but without some unwanted behaviours.
  \item We produce a CSP model of a 2-thread synchronisation object and show how multiple 2-thread synchronisation objects can be organised into an arbitrary $n$-thread barrier synchronisation object. We then use the CSP model to prove the correctness of the barrier synchronisation object.
\end{itemize}



% General correctness importance
% Concurrency - Therac \cite{BugSnag}
%Types of testing
%Why formal verification
%Contributions